Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-27 Manuscript/01-Intro.Rmd
Revision

Nonparametric regression models are appropriate alternatives to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be modeled with the summation of smoothing functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. The smoothing functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

>>>>

Nonparametric regression models are appropriate alternatives to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution, denoted by $EF(\mu, \phi)$, with mean $\mu$ and dispersion $\phi$, can be modeled with the summation of smoothing functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. The smoothing functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

### 2021-09-27 Manuscript/SS_GAM.tex
Revision

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be modeled with the summation of smoothing
functions, \(B_j(\cdot), j = 1, \dots, p\), of a given \(p\)-dimensional
vector of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

@@

\hypertarget{bayesian-hierarchical-additive-models}{%
\section{Bayesian Hierarchical Additive
Models}\label{bayesian-hierarchical-additive-models}}

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(\boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f(y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function. However, the direct integration of
smoothing penalty with sparsity penalty is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix \(\boldsymbol{S}_j\) is symmetric and positive
semi-definite for univariate smoothing functions, we apply
eigendecomposition on the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the non-linear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of non-linear
space.

>>>>

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution, denoted by \(EF(\mu, \phi)\), with mean
\(\mu\) and dispersion \(\phi\), can be modeled with the summation of
smoothing functions, \(B_j(\cdot), j = 1, \dots, p\), of a given
\(p\)-dimensional vector of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

@@

\hypertarget{bayesian-hierarchical-additive-models-bhams}{%
\section{Bayesian Hierarchical Additive Models
(BHAMs)}\label{bayesian-hierarchical-additive-models-bhams}}

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function. However, the direct integration of
smoothing penalty with sparsity penalty is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix \(\boldsymbol{S}_j\) is symmetric and positive
semi-definite for univariate smoothing functions, we apply
eigendecomposition on the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the non-linear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of non-linear
space.
